### 2013-05-31 doc/nips2013/factorial.tex
done with appendix

So far, we have relied on the existence of a bottleneck hidden variable $h_j$ on which
we have three conditionally independent views $x_{j_1},x_{j_2},x_{j_3}$.  Such bottlenecks do not always
exist; for example, in \reffig{factorialModels}(a), neither $h_1$ nor $h_2$ is
a bottleneck.
In this section, we present a technique to handle such models.
We start by creating a composite latent variable $c$ which is in
one-to-one correspondence with the existing latent variables $(h_1,h_2)$.
We can invoke $\TensorFactorize$ to obtain composite marginals and
conditional means, but the identities of $h_1$ and $h_2$ becomes lost in the
composite.  We develop a new algorithm $\UnshuffleFactorize$
to restore the identities and place constraints on the original
parameters.

\Fig{figures/factorialModels}{0.3}{factorialModels}{The factorial mixture model (a)
does not have three-view bottlenecks, so we first identify composite bottlenecks,
and then perform an \emph{unshuffling factorization} (b):
Given known composite conditional means
$L_c = \log \BP[x_j \mid c] \in \R^\ell$,
which are derived from combinations of unknown parameters
$f_a = \theta_1(\cdot, a), g_b = \theta_2(\cdot, b)$ and log-normalization scalars $Z_{ab}$,
we group difference vectors into bins (dashed boxes), and maximal bins into sources (red rounded boxes),
producing a linear program with variables $\{f_a\},\{g_b\},\{Z_{ab}\}$ and
a constraint for each $B_c$.
}

We focus on the restricted Boltzmann machine
\cite{smolensky86rbm,salakhutdinov09softmax},
which we refer to as an undirected factorial mixture model.
Let $h = (h_1, \dots, h_s) \in [\nh]^s$ be the latent variables (sources)
and $x = (x_1, x_2, x_3) \in [\nphix]^3$ be the observed variables.
The joint model probability is as follows:
\begin{align}
%p_\theta(x, h) = \prod_{i=1}^s p_\theta(h_i) \prod_{j=1}^3 p_\theta(x_j \mid h), \quad
%p_\theta(x_j \mid h) = \exp\left\{ \sum_{i=1}^s \theta_i(x_j, h_i) - A(\theta; h) \right\},
p_\theta(x, h) = \exp \left\{ \sum_{j=1}^3 \sum_{i=1}^s \theta_i(x_j, h_i) - A(\theta) \right\},
\end{align}
where each $\theta_i \in \R^{\nphix \times \nh}$ is the parameter matrix for the $i$-th source.

@@

Note that there is a combinatorial dimension to this problem since the association between
columns of $L$ and combinations of $\{ v_{ia} \}$ is unknown.
For example, example (\reffig{factorialModels}(b)),
consider $s=2$ sources, each taking on $k=3$ values
(for convenience, let $f_a = v_{1a}$ and $g_b = v_{2b}$).
The following algorithm, $\UnshuffleFactorize$,
creates a linear program that places constraints on
variables $\{v_{ia}\}$ and $Z_{\ba}$:
\begin{enumerate}

>>>>

So far, we have relied on the existence of a bottleneck hidden variable $h_j$ on which
we have three conditionally independent views $x_{j_1},x_{j_2},x_{j_3}$.  Such bottlenecks do not always
exist; for example, in \reffig{factorialMixtureModel}(a), neither $h_1$ nor $h_2$ is
a bottleneck.
In this section, we present a technique to handle such models.
We start by creating a composite latent variable $c$ which is in
one-to-one correspondence with the existing latent variables $(h_1,h_2)$.
We can invoke $\TensorFactorize$ to obtain composite marginals and
conditional means, but the identities of $h_1$ and $h_2$ becomes lost in the
composite.  We develop a new algorithm $\UnshuffleFactorize$
to restore the identities and place constraints on the original
parameters.

\Fig{figures/factorialMixtureModel}{0.3}{factorialMixtureModel}{The factorial mixture model (a)
does not have three-view bottlenecks, so we first identify composite bottlenecks,
and then perform an \emph{unshuffling factorization} (b):
Given known composite conditional means
$L_c = \log \BP[x_j \mid c] \in \R^\ell$,
which are derived from combinations of unknown parameters
$f_a = \theta_1(\cdot, a), g_b = \theta_2(\cdot, b)$ and log-normalization scalars $Z_{ab}$,
we group difference vectors into bins (dashed boxes), and maximal bins into sources (red rounded boxes),
producing a linear program with variables $\{f_a\},\{g_b\},\{Z_{ab}\}$ and
a constraint for each $B_c$.
}

We focus on the restricted Boltzmann machine
\cite{smolensky86rbm,salakhutdinov09softmax},
which we refer to as an undirected factorial mixture model
(the directed version is a two-layer sigmoid belief network \cite{saul96sigmoid}).
Let $h = (h_1, \dots, h_s) \in [\nh]^s$ be the latent variables (sources)
and $x = (x_1, x_2, x_3) \in [\nphix]^3$ be the observed variables.
The joint model probability is as follows:
\begin{align}
%p_\theta(x, h) = \prod_{i=1}^s p_\theta(h_i) \prod_{j=1}^3 p_\theta(x_j \mid h), \quad
%p_\theta(x_j \mid h) = \exp\left\{ \sum_{i=1}^s \theta_i(x_j, h_i) - A(\theta; h) \right\},
p_\theta(x, h) = \exp \left\{ \sum_{j=1}^3 \sum_{i=1}^s \theta_i(x_j, h_i) - A(\theta) \right\},
\end{align}
where each $\theta_i \in \R^{\nphix \times \nh}$ is the parameter matrix for the $i$-th source.

@@

Note that there is a combinatorial dimension to this problem since the association between
columns of $L$ and combinations of $\{ v_{ia} \}$ is unknown.
For example, example (\reffig{factorialMixtureModel}(b)),
consider $s=2$ sources, each taking on $k=3$ values
(for convenience, let $f_a = v_{1a}$ and $g_b = v_{2b}$).
The following algorithm, $\UnshuffleFactorize$,
creates a linear program that places constraints on
variables $\{v_{ia}\}$ and $Z_{\ba}$:
\begin{enumerate}
